Fractional Decomposition Tree Algorithm: A tool for studying the integrality gap of Integer Programs

### 2020-08-02 FDT2020/Introduction.tex
More addition of careful definitions, removal of redundancy, etc.

\section{Introduction}
In this paper we focus on finding feasible solutions to binary Integer Linear Programs (IP). Informally, an integer program is the optimization of a linear objective function subject to linear constraints, where the variables must take integer values. Binary variables represent yes/no decisions. Integer Programming (and more generally Mixed Integer Linear Programming (MILP)) can model many practical optimization problems including scheduling, logistics and resource allocation. It is  NP-hard even to determine if an IP instance has a feasible solution~\cite{GareyJohnson}. Given its practical importance, however, there are many commercial (e.g. CPLEX, GUROBI, XPRESS) and free (e.g. CBC) solvers that for specific IP instances can often find solutions that are optimal within a given tolerance. Still we have formulated moderate-sized IP instances that no commercial solver can currently solve. Thus there is value in heuristics to find feasible solutions for general IP instances (see e.g.~\cite{HanafiT2017}).

@@

Many researchers (see \cite{davids,vazirani}) have developed polynomial time LP-based approximation algorithms that find solutions for special classes of IPs whose cost are provably smaller than $C\cdot z_{LP}(I,c)$. If the analysis uses the LP bound to prove the approximation quality, then $C$ is at least the integrality gap.
%However, for many combinatorial optimization problems there is a limit to such techniques based on LP relaxations, represented  by the {integrality gap} of the IP formulation.
\begin{definition}
The integrality gap $g(I)$ for instance $I$ is: $$g(I)= \max_{c\geq 0}\frac{z_{IP}(I,c)}{z_{LP}(I,c)},$$
where $z_{IP}(I,c)$ is the optimal solution to the integer program and $z_{LP}(I,c)$ is the solution to the linear-programming relaxation.
\end{definition}
\cindy{I'm not sure I got an answer for this, so I'll just make it clear this is the definition we use: Does the integrality gap always require positive objective coefficients?  Or is that an extra restriction in our definition?  Is there a reference that has that restriction?} 
For example consider the minimum cost 2-edge-connected multi-subgraph problem (2EC): Given a graph $G=(V,E)$ and $c\in \mathbb{R}^E_{\geq 0}$, 2EC asks for the minimum cost 2-edge-connected subgraph of $G$, with multiedges allowed. A graph is 2-edge-connected if there are at least two edge-disjoint paths between every pair of vertices.  A linear programming relaxation for this problem known as the subtour elimination relaxation  is
 \begin{equation}\min \{cx: \sum_{e\in \delta(U)}x_e \geq 2 \mbox{ for } \emptyset \subsetneq U \subsetneq V,\; x\in [0,2]^{E}\}, 
 \end{equation}
where $\delta(U)$ for vertex subset $U$ is the set of edges that cross the cut defined by $U$.  That is, each $e \in \delta(U)$ has one endpoint in $U$ and the other endpoint in $V-U$.
 In this case the instance-specific integrality gap is the integrality gap of the subtour-elimination relaxation for the 2EC on graph with $n$ vertices. Alexander et al. \cite{alexander2006integrality} showed this instance-specific integrality gap
%of the subtour elimination relaxation for the 2EC
is at most $\frac{7}{6}$ for instances with $n= 10$ .

@@

We give a general approximation framework for solving binary IPs.
Consider the set of points described by sets $S(I)$ and $P(I)$ as in (\ref{S}) and (\ref{P}), respectively. Assume in addition that $S(I),P(I)\subseteq [0,1]^n$.
\cindy{Is it $S(I) \subseteq \{0,1\}^n$}
For a vector $x\in \mathbb{R}_{\geq 0}^n$ such that $x\in P(I)$, let the {\em support} of $x$ be $\spp(x)= \{i \in [n]: x_i \neq 0\}$. For an integer $\beta$ let $\{\beta\}^n$ be the vector $y\in \mathbb{R}^n$ with $y_i=\beta$ for $i\in [n]$.

We introduce the \textit{Fractional Decomposition Tree Algorithm} (FDT) which runs in polynomial time algorithm. Given a point $x\in P(I)$ FDT produces a convex combination of feasible points in $S(I)$ that are dominated by a ``factor" $C$ of $x$ in the coordinates corresponding to $x$. If $C = g(I)$, it would be optimal. However we can only guarantee a factor of $g(I)^{|\spp(x)|}$. FDT relies on iteratively solving linear programs that are about the same size as the description of $P(I)$.

@@

For a generic IP instance $I$ it is NP-hard to even decide if the set of feasible solutions $S(I)$ is empty or not. There are a number of heuristics for this purpose, such as the feasibility pump algorithm \cite{fp1,fp2}. These heuristics are often very effective and fast in practice, however, they can sometimes fail to find a feasible solution. Moreover, these heuristics do not provide any bounds on the quality of the solution they find.

Here is how the FDT algorithm works in a high level: in iteration $i$ the algorithm maintains a convex combination of  vectors in $\mathcal{D}(L(I))$ that have a 0 or 1 value for coordinates indexed $0,\ldots,i-1$. Let $y$ be a vector in the convex combination in iteration $i$ of the algorithm. We solve a linear programming problem that gives us $\theta\in [0,1]$ and $y^0,y^1\in \mathcal{D}(L(I))$ such that $g(I) y\geq \theta_1y^0 + (1-\theta) y^1$ and $y^0_i=0$ and $y^1_i=1$. We then replace $y$ in the convex combination with $\frac{\theta}{g(I)}y^0 +\frac{1-\theta}{g(I)}y^1$. Repeating this for every vector in the convex combination from previous iteration yields a convex combination  of points that is ``more'' integral. If in any iteration there are too many points in the convex combination we solve a linear programming problem that ``prunes'' the convex combination. At the end we find a convex combination of integer solutions $\mathcal{D}(L(I))$. For each such solution $z$ we invoke the DomToIP algorithm (see Section \ref{domTOIP}) to find $z'\in S(I)$ where $z'\leq z$.

One can extend the FDT algorithm for binary IPs into covering $\{0,1,2\}$ IPs by losing a factor $2^{|\spp(x)|}$ on top of the loss for FDT. In order to eradicate this extra factor, we need to treat the coordinate $i$ with $x_i=1$ differently. We focus on the \textsc{2-edge-connected multi-subgraph graph problem (2EC)}: Given a graph $G=(V,E)$ and $c\in \mathbb{R}^{E}_{\geq 0}$ find a 2-edge-connected multi-subgraph of $G$ with minimum cost. The natural linear programming relaxation for this problem is 
\begin{equation}
\min \{cx \; : \; x(\delta(U))\geq 2 \; \text{ for }\emptyset \subset U \subset V, \; x\in [0,2]^E\}
\end{equation}
We denote the feasible region of this LP by $\subtour(G)$. Let $\2ec(G)$ be the convex hull of incidence vectors of 2-edge-connected multi-subgraphs of graph $G$. Following the definition in (\ref{gapproblem}) have
\begin{equation}
g(\2ec) = \max_{c\geq 0 , G}\frac{\min_{x\in \2ec(G)} cx}{\min_{x\in \subtour(G)} cx}.
\end{equation}

@@

\subsection{Experiments.} Although the bound guaranteed in both Theorems \ref{binaryFDT} and \ref{FDT2EC} are very large, we show that in practice, the algorithm works very well for network design problems described above. We show how one might use FDT to investigate the integrality gap for such well-studied problems. \cindy{TODO: forward pointers to the sections with more details.}

@@

In the \textsc{minimum vetex cover problem (VC)} we are given a graph $G=(V,E)$ and $c\in \mathbb{R}^E_{\geq 0}$. A subset of $U$ of $V$ is a \textit{vertex cover} if for $e\in E$ at least one endpoint of $e$ is in $U$. The goal in VC is to find the minimum cost vertex cover. The linear programming relaxation for VC is
\begin{equation}
\min \{cx \; : \; x_u + x_v \geq	 1 \text{ for } e=uv \in E, \; x\in [0,1]^{V}\}.
\end{equation}
The integrality gap of this formulation is exactly 2 \cite{davids}. Austrin, Khot and Safra~\cite{UGhardVC} show that it is UG-hard to approximte VC within any factor sctrictly better than 2. We compare  FDT and the feasbility pump heuristic \cite{fp1} on the small instances of PACE\footnote{ Parameterized Algorithms and Computational Experiments: \url{https://pacechallenge.org/2019/}} 2019 challenge test cases \cite{PACE}. 
\subsubsection{Tree augmentation problem}
In the \textsc{Tree Augmentation Problem (TAP)} we are given a  graph $G=(V,E)$, a spanning tree $T$ of $G$. We also have a cost vector $c\in \mathbb{R}^{E\setminus T}_{\geq 0}$. A subset $F$ of $E\setminus T$ is called a \textit{feasible augmentation} if $(V,T\cup F)$ is a 2-edge-connected graph. In TAP we seek the minimum cost feasible augmentation. The natural linear programming relaxation for TAP is 
\begin{equation}
\min \{cx\; : \; \sum_{\ell \in \cov(e)} x_{\ell} \geq 1 \text{ for } e\in T, \; x\in [0,1]^{E\setminus T}\}.
\end{equation}
where $\cov(e)$ is set of edges $\ell \in E\setminus T$ such that $e$ is in the unique cycle of $T\cup \{\ell\}$. We call the LP above the cut-LP. The integrality gap of the cut-LP is known to be between $\frac{3}{2}$ \cite{32gaptap} and $2$ \cite{FJ81}. We create random fractional extreme points of the cut-LP and apply FDT to find integral solutions. For the instances that we create, the blow-up factor is always below $\frac{3}{2}$ providing an upper bound for such instances.

>>>>

\section{Introduction}
In this paper we focus on finding feasible solutions to binary Integer Linear Programs (IP). Informally, an integer program is the optimization of a linear objective function subject to linear constraints, where the variables must take integer values. Binary variables represent yes/no decisions. Integer Programming (and more generally Mixed Integer Linear Programming (MILP)) can model many practical optimization problems including scheduling, logistics and resource allocation.

It is  NP-hard even to determine if an IP instance has a feasible solution~\cite{GareyJohnson}. Given its practical importance, however, there are many commercial (e.g. CPLEX, GUROBI, XPRESS) and free (e.g. CBC) solvers that for specific IP instances can often find solutions that are optimal within a given tolerance. Still we have formulated moderate-sized IP instances that no commercial solver can currently solve. Thus there is value in heuristics to find feasible solutions for general IP instances (see e.g.~\cite{HanafiT2017}).
%For a generic IP instance $I$ it is NP-hard to even decide if the set of feasible solutions $S(I)$ is empty or not.
%There are a number of heuristics for this purpose, such as the
These heuristics, such as the popular Feasibility pump algorithm \cite{fp1,fp2}, are often effective and fast in practice. However, the heuristics can sometimes fail to find a feasible solution. Moreover, these heuristics do not provide any bounds on the quality of the solution they find.

@@

Many researchers (see \cite{davids,vazirani}) have developed polynomial time LP-based approximation algorithms that find solutions for special classes of IPs whose cost are provably smaller than $C\cdot z_{LP}(I,c)$. If the analysis uses the LP bound to prove the approximation quality, then $C$ is at least the integrality gap.
%However, for many combinatorial optimization problems there is a limit to such techniques based on LP relaxations, represented  by the {integrality gap} of the IP formulation.
\cindy{Arash: Just read and comment it out if you're OK with this. I'm not sure I got an answer for the following questions, so I'll just make it clear this is the definition we use: Does the integrality gap always require positive objective coefficients?  Or is that an extra restriction in our definition?  Is there a reference that has that restriction?} 
\begin{definition}
The integrality gap $g(I)$ for instance $I$ is: $$g(I)= \max_{c\geq 0}\frac{z_{IP}(I,c)}{z_{LP}(I,c)},$$
where $z_{IP}(I,c)$ is the optimal solution to the integer program and $z_{LP}(I,c)$ is the solution to the linear-programming relaxation.
\end{definition}

For example consider the minimum cost \textsc{2-edge-connected multi-subgraph problem (2EC)}: Given a graph $G=(V,E)$ and $c\in \mathbb{R}^E_{\geq 0}$, 2EC asks for the minimum cost 2-edge-connected subgraph of $G$, with multiedges allowed. A graph is 2-edge-connected if there are at least two edge-disjoint paths between every pair of vertices.  A linear programming relaxation for this problem, known as the {\em subtour-elimination} relaxation, is
 \begin{equation}\min \{cx: \sum_{e\in \delta(U)}x_e \geq 2 \mbox{ for } \emptyset \subsetneq U \subsetneq V,\; x\in [0,2]^{E}\}, 
 \label{eq:subtour}
 \end{equation}
where $\delta(U)$ for vertex subset $U$ is the set of edges that cross the cut defined by $U$.  That is, each $e \in \delta(U)$ has one endpoint in $U$ and the other endpoint in $V-U$.
 In this case the instance-specific integrality gap is the integrality gap of the subtour-elimination relaxation for the 2EC on graph with $n$ vertices. Alexander et al. \cite{alexander2006integrality} showed this instance-specific integrality gap
%of the subtour elimination relaxation for the 2EC
is at most $\frac{7}{6}$ for instances with $n= 10$ .

@@

We give a general approximation framework for solving binary IPs.
Consider the set of points described by sets $S(I)$ and $P(I)$ as in (\ref{S}) and (\ref{P}), respectively. Assume in addition that $S(I),P(I)\subseteq [0,1]^n$.
\cindy{Is it $S(I) \subseteq \{0,1\}^n$?}
For a vector $x\in \mathbb{R}_{\geq 0}^n$ such that $x\in P(I)$, let the {\em support} of $x$ be $\spp(x)= \{i \in [n]: x_i \neq 0\}$. For an integer $\beta$ let $\{\beta\}^n$ be the vector $y\in \mathbb{R}^n$ with $y_i=\beta$ for $i\in [n]$.

We introduce the \textit{Fractional Decomposition Tree Algorithm} (FDT) which runs in polynomial time algorithm. Given a point $x\in P(I)$ FDT produces a convex combination of feasible points in $S(I)$ that are dominated coordinatewise by a ``factor" $C$ times $x$.
% in the coordinates corresponding to $x$.
If $C = g(I)$, it would be optimal. However we can only guarantee a factor of $g(I)^{|\spp(x)|}$. FDT iteratively solves linear programs that are about the same size as the description of $P(I)$.

@@

\cindy{Arash, please check how I defined $L(I)$.  It was not defined before.}
\cindy{TODO: This implies we know $g(I)$, which is not true in general.  We should add how we bound/estimate it.  I'll return to this after putting that detail back in core if you (Arash) haven't added it by then.}
Here is how the FDT algorithm works at a high level for an instance $I$ with LP feasible region $L(I)$: in iteration $i$ the algorithm maintains a convex combination of  vectors in $\mathcal{D}(L(I))$ that have a 0 or 1 value for coordinates indexed $0,\ldots,i-1$. Let $y$ be a vector in the convex combination from iteration $i-1$. We solve a linear programming problem that gives us $\theta\in [0,1]$ and $y^0,y^1\in \mathcal{D}(L(I))$ such that $g(I) y\geq \theta y^0 + (1-\theta) y^1$ and $y^0_i=0$ and $y^1_i=1$. We then replace $y$ in the convex combination with $\frac{\theta}{g(I)}y^0 +\frac{1-\theta}{g(I)}y^1$. Repeating this for every vector in the convex combination from the previous iteration yields a new convex combination of points. It is ``more'' integral because now all vectors in the convex combination are integral in their first $i+1$ elements. If in any iteration there are too many points in the convex combination,  we solve a linear programming problem that ``prunes'' the convex combination. At the end we have a convex combination of integer solutions $\mathcal{D}(L(I))$. For each such solution $z$ we invoke the DomToIP algorithm (see Section \ref{domTOIP}) to find $z'\in S(I)$ where $z'\leq z$.

One can extend the FDT algorithm for binary IPs into covering\footnote{A covering IP has nonnegative constraint matrix, objective coefficients and right-hand side ($A,b,c$).} $\{0,1,2\}$ IPs by losing a factor $2^{|\spp(x)|}$ on top of the loss for FDT. To eradicate this extra factor, we must treat the coordinate $i$ with $x_i=1$ differently. We focus on the 2-edge-connected multi-subgraph graph problem (2EC) defined above.  It's subtour-elimination LP relaxation is given in (\ref{eq:subtour}).
%}: Given a graph $G=(V,E)$ and $c\in \mathbb{R}^{E}_{\geq 0}$ find a 2-edge-connected multi-subgraph of $G$ with minimum cost. The natural linear programming relaxation for this problem is 
%\begin{equation}
%\min \{cx \; : \; x(\delta(U))\geq 2 \; \text{ for }\emptyset \subset U \subset V, \; x\in [0,2]^E\}
%\end{equation}
For input graph $G$, let $\subtour(G)$ denote the feasible region of this LP. Let $\2ec(G)$ be the convex hull of incidence vectors\footnote{The incidence vector $x$ of a 2-edge-connected multigraph has an element for each edge $e$, with $x_e \in \{0,1,2\}$ indicating the number of times edge $e$ appears in the solution.} of 2-edge-connected multi-subgraphs of graph $G$. Following the definition in (\ref{gapproblem}) have
\begin{equation}
g(\2ec) = \max_{c\geq 0 , G}\frac{\min_{x\in \2ec(G)} cx}{\min_{x\in \subtour(G)} cx}.
\end{equation}

@@

\subsection{Experiments.} Although the bounds guaranteed in both Theorems \ref{binaryFDT} and \ref{FDT2EC} are large, in Section~\ref{sec:experiment} we show that in practice, the algorithm works well for network design problems like those described above. We also show how one might use FDT to investigate the integrality gap for such well-studied problems.
%\cindy{TODO: forward pointers to the sections with more details.}

@@

In the \textsc{minimum vertex cover problem (VC)} we are given a graph $G=(V,E)$ and $c\in \mathbb{R}^E_{\geq 0}$. A subset of $U$ of $V$ is a \textit{vertex cover} if for all $e\in E$ at least one endpoint of $e$ is in $U$. The goal in VC is to find the minimum-cost vertex cover. The linear programming relaxation for VC is
\begin{equation}
\min \{cx \; : \; x_u + x_v \geq	 1 \text{ for all } e=uv \in E, \; x\in [0,1]^{V}\}.
\end{equation}
The integrality gap of this formulation is exactly 2 \cite{davids}. Austrin, Khot and Safra~\cite{UGhardVC} show that it is UG-hard to approximate VC within any factor strictly better than 2. We compare  FDT and the feasbility pump heuristic \cite{fp1} on the small instances of PACE\footnote{ Parameterized Algorithms and Computational Experiments: \url{https://pacechallenge.org/2019/}} 2019 challenge test cases \cite{PACE}. \cindy{TODO: summary of results.  Something like FDT was 3-5x slower, but still not too long, and almost always had a much lower gap.}
\subsubsection{Tree augmentation problem}
In the \textsc{Tree Augmentation Problem (TAP)} we are given a  graph $G=(V,E)$, a spanning tree $T$ of $G$. We also have a cost vector $c\in \mathbb{R}^{E\setminus T}_{\geq 0}$. A subset $F$ of $E\setminus T$ is called a \textit{feasible augmentation} if $(V,T\cup F)$ is a 2-edge-connected graph. In TAP we seek the minimum cost feasible augmentation. The natural linear programming relaxation for TAP is 
\begin{equation}
\min \{cx\; : \; \sum_{\ell \in \cov(e)} x_{\ell} \geq 1 \text{ for } e\in T, \; x\in [0,1]^{E\setminus T}\}.
\end{equation}
where $\cov(e)$ is set of edges $\ell \in E\setminus T$ such that $e$ is in the unique cycle of $T\cup \{\ell\}$. We call the LP above the cut-LP. The integrality gap of the cut-LP is known to be between $\frac{3}{2}$ \cite{32gaptap} and $2$ \cite{FJ81}. We create random fractional extreme points of the cut-LP and apply FDT to find integral solutions. For the instances that we create, the blow-up factor is always below $\frac{3}{2}$ providing an upper bound for such instances.
